---
title: "Hartigan-Wong Experiments"
output: html_notebook
---
## Purpose
Investigate the performance of the Hartigan-Wong algorithm for the *k*-means problems used in the preprint [Breathing-K-Means](https://arxiv.org/abs/2006.15666).

### Usage
- run whole notebook in [RStudio](https://www.rstudio.com/) for statistics on 2D or High-D data sets
- select either 2D problems or High-D problems in the "Problem group" section below
- graphically display single problems using function ```run_and_plot()```

```{r}
library(data.table)
library(tidyverse)
```
### Simulation parameters
```{r}
repetitions=10    # runs per problem (), how often should each problem be run
                  # this is *not* the Parameter nstart from kmeans (which is set to 10 in the code below)
n_problems=20     # (select, e.g., 5,  to run only on the first 5 problems from the lists below)
```
### Problem group (2-D or High-D)
```{r}
probs=probs2D     # uncomment to run two-dimensional problems
#probs=probsHighD # uncomment to run high-dimensional problems
```
### 2D problems
```{r}
probs2D = list(
  'Jain#30',
  'R15#30',
  'D31#100',
  'Aggregation#200',
  'Flame#80',
  'Spiral#80',
  'S2#100',
  'Unbalance#100',
  'Birch1#100',
  'Birch3#100',
  'gmd5x5#50',
  'gmd8x8#128',
  'noise#100',
  'vardensity#100',
  'uniform#100',
  'sq6x6-2x2#144',
  'randangles#60',
  'stains#100',
  'correlated#100',
  'rice#120'
)

```
### High-D problems
```{r}
probsHighD = list(
  'Cloud#25','Cloud#50','Cloud#100','Cloud#150','Cloud#200',
  'Norm25-new#25','Norm25-new#50','Norm25-new#100','Norm25-new#150','Norm25-new#200',
  'Intrusion#25','Intrusion#50','Intrusion#100','Intrusion#150','Intrusion#200',
  'Spam#25','Spam#50','Spam#100','Spam#150','Spam#200'
  )
```
### Function to read problem data
```{r}
read_data = function(problem){
  # returns name, k, data
  L <- as.list(strsplit(problem, "#")[[1]])
  names(L)=c("name","k")
  if (problem %in% probsHighD) {
    extension = ".zip"
  } else {
    extension = ".csv"
  }
  path_to_data = "../data/" # modify as required
  filename = paste(path_to_data,L$name,extension,sep="")
  X <- read_csv(file = filename, col_names = FALSE, col_types = cols())
  return( list(name=L$name, k=strtoi(L$k), X=X))
}
```

### Function to run Hartigan-Wong
(with 10 starts as specified by the nstart parameter)
```{r}
run_hartigan_wong = function(X, k) {
  # returns SSE (phi), warning, cputime
  # catches errors and warnings
  t0=Sys.time()
  tryCatch({
    cl <-
      kmeans(
        X,
        centers = k,
        iter.max = 100,
        nstart = 10,
        algorithm = "Hartigan-Wong"
      )
    mywarn = "ok"
  }, error = function(e) {
    print(e)
    mywarn <<- toString(e)
  },
  warning = function(e) {
    print(e)
    mywarn <<- toString(e)
  })
  t=Sys.time()-t0
  return(list(phi=cl$tot.withinss,warn=mywarn,time=t,cl=cl))
}
```

### Repeated runs on a list of problems
```{r}
# perform reps" runs for each problem
make_stat = function(reps=1,probs) {
  name=c()
  k=c()
  err=c()
  time=c()
  message=c()
  for (p in probs){
    print(p)
    # read data
    L = read_data(p)
    for (z in 1:reps){
      R=run_hartigan_wong(L$X,L$k)
      name=c(name,L$name)
      k=c(k,L$k)
      err=c(err,R$phi)  
      time=c(time,R$time)
      message=c(message,R$warn)
    }
  }
  data.table(name,k,err,time,message)
}
```

## Make mass runs
```{r}
df=make_stat(reps=repetitions,probs[1:n_problems])
```

### Show raw results
```{r}
raw_table=as_tibble(df)
raw_table
```

### Summarize by *k*-means problem
Column "success" indicates the number of non-failed runs
```{r}
make_summary = function() {
  #
  # final evaluation
  # expects result rows in df
  #
  res = raw_table  %>% 
    filter(message=="ok") %>%  # only select rows with successful runs
    group_by(name,k) %>%  # group by data set 
    summarise(err=mean(err),t=mean(time),success=n(),.groups = 'drop') %>% # mean
    mutate(err = formatC(err,format="e",digits=3), t=formatC(t,format="f",digits=5)) # format
  res
}
final = make_summary()

final
```
### Show failed runs summary
```{r}
raw_table %>% filter(message != "ok") %>% group_by(name,k) %>% summarise(failed=n(), .groups = 'drop')
```
## Failed runs details
```{r}
raw_table %>% filter(message!="ok") %>% filter(name!="xxx") %>% mutate(err=as.numeric(err))
```
## Run Hartigan-Wong algorithm and plot result
```{r}
run_and_plot = function(p) {
  # get data
  L = read_data(p)
  # run Hartigan-Wong
  R = run_hartigan_wong(L$X,L$k)
  too_few = FALSE
  # evaluate
  if (R$warn != "ok") {
    print("Hartigan-Wong algorithm failed")
    print(R$warn)
    if (nrow(R$cl$centers) < L$k){
      print(paste("delivered only",as.character(nrow(R$cl$centers)),"of",as.character(L$k),"centers."))
      print(R$cl$centers)
      too_few = TRUE
    } else {
      print(paste("delivered all",as.character(nrow(R$cl$centers)),"of",as.character(L$k),"centers."))      
    }
  } 
  
  sse = formatC(R$phi, format = "e", digits = 3)
  X=L$X
  colnames(X) <- c("x", "y")
  n=nrow(X)
  if (n<1000) {
    cex=0.5
  } else if (n < 5000) {
    cex=0.3
  } else {
    cex=0.1
  }
  p <- plot(
    X[c("x","y")],
    col = "green",
    asp = 1,
    cex = cex,
    pch = 16,
    if (too_few) {
      main = paste(p,", failed: only ",as.character(nrow(R$cl$centers))," of ",
                   as.character(L$k), " centroids delivered (a.o.t.)", sep="")  
    } else {
      main = paste(p,", ","  SSE = ",sse,sep="")
    }
  )
  points(R$cl$centers, col = 1, pch = 2,cex=1.0)
  grid()
}
```
# Some examples solutions
```{r}
for (p in list("Jain#30","Flame#80","randangles#60","randangles#20","gmd5x5#50")) {
    run_and_plot(p)
  }
```
```{r}
run_and_plot("Birch1#100")
```
## Problem Norm25-new#25
25 clusters and 25 centroids. Only 2 from 15 dimensions shown. A good solution needs to position one centroid in each cluster.
```{r}
run_and_plot("Norm25-new#25")
```

